The Impact of Alternation

### 2017-05-07 draft.tex
switched to IEEE Trans

\documentclass{llncs}

@@

\author{Radu Iosif \and Xiao Xu}

\institute{Verimag/CNRS/Universit\'e de Grenoble Alpes}

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclclclcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} && 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ is not an atom} \\
\overline{\phi} & \equiv & \phi \text{  if $\phi \in Q$}\hspace*{1cm} &&
\overline{\phi} & \equiv & \neg\phi \text{  if $\phi \not\in Q$ is an atom}
\end{array}\]
It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
every formula $\phi \in \Form^+(Q,\vec{x})$. If $\phi$ does not
contain boolean variables, then $\overline{\phi}$ and $\neg\phi$ are
equivalent, and in general:

@@

\begin{lemma}\label{lemma:closure}
  For any automata $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
  and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$ such that $Q_1
  \cap Q_2 = \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$,
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \eqiff \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \eqiff \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{rcl}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iff & 
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee
  \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) \\
  & \iff & \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots
  a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{rcll}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & 
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \text{ by induction hypothesis} \\
  & \eqiff & \overline{\Delta(\iota_1,a_1\ldots,a_n)} & \text{ by the definition of $\overline{\phi}$ .}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be two valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the boolean operations, including complementation, can be built in
linear time. This is not very surprising, since linear time
intersection and complementation are also possible for alternating
automata over finite alphabets \cite{ChandraKozenStockmeyer81}.

@@

Consequently, we give up on the guarantee for termination and develop
semi-algorithmic methods that are supposed to be
both: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$), and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

@@

The solution we adopt is the use of \emph{abstraction} to ensure
termination of invariant computations. However, it is worth pointing
out from the start that abstraction will allow us to build invariants
which are not necessarily safety invariants. To meet the latter
condition, we resort to \emph{counterexample guided abstraction
  refinement} (CEGAR).

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can occur negated in some $\phi \in \preds$. We assume that $\preds$
is fixed for now, and refer to these formulae as \emph{predicates}.
Given a formula $\phi$, we denote by $\abs{\phi} \equiv \bigwedge
\set{\pi \in \preds \mid \phi \models \pi}$ the abstraction of $\phi$
w.r.t. the predicates in $\preds$. The abstract version of the
post-image and acceptance condition are defined as follows:
{\footnotesize\[ \AbsPost{\A}(\phi,\varepsilon) \equiv \phi
\hspace*{2mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} 
\hspace*{2mm}
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)\]}
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \eqiff
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{equation}\label{eq:interpolation-problem}
\pi(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1} \cup
Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)
\end{equation} 
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, $k\in[1,n]$ uses
\emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in [0,n]$,
which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$, 
%
\item for each $\ell\in[1,n]$: $\theta_\ell \equiv
  \bigwedge_{q_{\ell-1}\in R_{\ell-1}} (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$.
\end{compactitem}
Finally, we set $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus
  F} (q_n \rightarrow \bot)$. The intuition is that $\theta_0, \ldots,
\theta_n$ simulate the run of $\A$ over $u$ and $\theta_{n+1}$ is the
acceptance condition, which forces the last remaining non-final states
to be false.

@@

  (\ref{it1:progress}) Assume that $\pi(u)$ is satisfiable and let
  $\beta_0,\ldots,\beta_n$ and $\nu_0,\ldots,\nu_n$ be boolean and
  data valuations for $Q_0,\ldots,Q_n$ and
  $\vec{x}_0,\ldots,\vec{x}_n$, respectively, such that
  $\bigcup_{i=0}^n \beta_i \cup \bigcup_{i=0}^n \nu_i \models
  \pi(u)$. We build a model for $\Accept{\A}(u) \equiv
  \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$. Let
  $\phi_0,\phi_1,\ldots,\phi_n$ be the run of $\A$ over $u$. We show
  that $\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$ has a model, by
  induction on $n\geq0$. For the base case $n=0$, $\phi_0 \equiv \iota
  \equiv \theta_0[Q/Q_0]$ has the model $\beta(q) = \beta_0(q_0)$, for
  all $q \in Q$. For the induction step $n > 0$, we assume that
  $\phi_{n-1}(Q,\vec{x}_0\cup\ldots\vec{x}_{n-1})$ has the model
  $\beta \cup \nu_0 \cup \ldots \nu_{n-1}$, where $\beta(q) =
  \beta_{n-1}(q_{n-1})$, for all $q \in Q$. By the definition of
  $\pi(u)$, we have: \[\theta_n \equiv \bigwedge_{q_{n-1} \in
    \fv{\Bool}{\theta_{n-1}} \cap Q_{n-1}} \left(q_{n-1} \rightarrow
  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]\right)\]

@@

  (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
    interpolant for $\pi(u)$, the following entailments hold: \begin{compactitem}
    \item $\theta_0 \models I_1[Q_0/Q]$, 
      %
    \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, for all $k\in[1,n]$. 
      %
    \item $I_{n+1}[Q_{n+1}/Q] \wedge \theta_{n+1} \models \bot$. 
    \end{compactitem}
    We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$
    by induction on $n \geq 0$. This is sufficient to conclude because
    $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
    a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
    \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we
    have $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv
    \theta_0[Q/Q_0] \models I_1$. For the induction step $n>0$, we
    compute:
    \[\begin{array}{rcll}
    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] & \equiv & \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] & \text{ def. of $\AbsPost{\A}$} \\
    & \models & \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n & \text{ ($\Asterisk$)} \\
    & \models & \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n & \text{ by ind. hyp.} \\ 
    & \models & I_n[Q_n/Q] & \hfill\text{\qed}
    \end{array}\]
    We prove the final point ($\Asterisk$). If
    $\beta_n\cup\nu_{n-1}\cup\nu_n \models
    \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q]$,
    for some valuations $\beta_{n-1} : Q_{n-1} \rightarrow \booli$ and
    $\nu_i : \vec{x}_i \rightarrow \Data^\I$, for $i=n-1,n$, then we
    build a valuation $\beta_{n-1}$ such that $\beta_{n-1}\cup\beta_n
    \cup \nu_{n-1}\cup\nu_n \models \AbsPost{\A}(\iota,a_1\ldots
    a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$, as follows. For each
    occurrence of a subformula $\Delta(q,a_n)$ in
    $\AbsPost{\A}(\iota,a_1\ldots a_{n-1})$ we set
    $\beta_{n-1}(q_{n-1}) = \true$ if $\beta_n\cup\nu_{n-1}\cup\nu_n
    \models \Delta(q,a_n)[Q_n/Q]$ and $\beta_{n-1}(q_{n-1}) = \false$,
    otherwise. Since there are no negated occurrences of such
    subformulae, the definition of $\beta_{n-1}$ is consistent, and
    the check $\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
    \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$
    is straightforward. \qed}

@@

\begin{algorithm}[t!]
\begin{algorithmic}[0]
\State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
over the alphabet $\Sigma$ of input events

\State {\bf output}: $\true$ if $L(\A)=\emptyset$ and a data word $w
\in L(\A)$ otherwise
\end{algorithmic}

\begin{algorithmic}[1]
  \State $\preds \leftarrow \set{\bot}$, $\worklist \leftarrow
  \tuple{\rootNode}$, $\Art = \tuple{N,E,\Lambda,\lhd}$, where $N \leftarrow
  \emptyset$, $E \leftarrow \emptyset$, $\Lambda \leftarrow
  \set{(\rootNode,\iota)}$ and $\lhd \leftarrow \emptyset$

>>>>

% \documentclass{llncs}
\documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}

@@

\author{Radu Iosif and Xiao Xu \\
Verimag/CNRS/Universit\'e de Grenoble Alpes\\
Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
\end{array}\]
It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
every formula $\phi \in \Form^+(Q,\vec{x})$. If $\phi$ does not
contain boolean variables, then $\overline{\phi}$ and $\neg\phi$ are
equivalent, and in general:

@@

\begin{lemma}\label{lemma:closure}
  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
  L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \eqiff \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \eqiff \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{ll}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \eqiff \\
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \eqiff \\
  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{lcl}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & \text{ by induction hypothesis} \\
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \eqiff & \text{ by the definition of $\overline{\phi}$} \\
  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the boolean operations, including complementation, can be built in
linear time. This is not surprising, as linear time intersection and
complementation are also possible for alternating automata over finite
alphabets \cite{ChandraKozenStockmeyer81}.

@@

Consequently, we give up on the guarantee for termination and develop
semi-algorithms that are supposed to be: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$), and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

@@

The solution we adopt is the use of abstraction to ensure termination
of invariant computations. However, it is worth pointing out from the
start that abstraction will allow us to build invariants which are not
necessarily safety invariants. To meet the latter condition, we resort
to counterexample guided abstraction refinement (CEGAR).

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can occur negated in some $\phi \in \preds$. We assume that $\preds$
is fixed for now, and refer to these formulae as \emph{predicates}.
Given a formula $\phi$, we denote by $\abs{\phi} \equiv \bigwedge
\set{\pi \in \preds \mid \phi \models \pi}$ the abstraction of $\phi$
w.r.t. the predicates in $\preds$. The abstract version of the
post-image and acceptance condition are defined as follows:
\[\begin{array}{llcl}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi & 
\AbsPost{\A}(\phi,ua) & \equiv & \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
&
\AbsAccept{\A}(u) & \equiv & \AbsPost{\A}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \eqiff
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{eqnarray}\label{eq:interpolation-problem}
\pi(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \\ 
&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
\end{eqnarray}
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, $k\in[1,n]$ uses
\emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in [0,n]$,
which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$, 
%
\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
  (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
  $\ell\in[1,n]$.
\end{compactitem}
Finally, we set $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus
  F} (q_n \rightarrow \bot)$. The intuition is that $\theta_0, \ldots,
\theta_n$ simulate the run of $\A$ over $u$ and $\theta_{n+1}$ is the
acceptance condition, which forces the last remaining non-final states
to be false.

@@

  (\ref{it1:progress}) Assume that $\pi(u)$ is satisfiable and let
  $\beta_0,\ldots,\beta_n$ and $\nu_0,\ldots,\nu_n$ be boolean and
  data valuations for $Q_0,\ldots,Q_n$ and
  $\vec{x}_0,\ldots,\vec{x}_n$, respectively, such that
  $\bigcup_{i=0}^n \beta_i \cup \bigcup_{i=0}^n \nu_i \models
  \pi(u)$. We build a model for $\Accept{\A}(u) \equiv
  \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$. Let
  $\phi_0,\phi_1,\ldots,\phi_n$ be the run of $\A$ over $u$. We show
  that $\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$ has a model, by
  induction on $n\geq0$. For the base case $n=0$, $\phi_0 \equiv \iota
  \equiv \theta_0[Q/Q_0]$ has the model $\beta(q) = \beta_0(q_0)$, for
  all $q \in Q$. For the induction step $n > 0$, we assume that
  $\phi_{n-1}(Q,\vec{x}_0\cup\ldots\vec{x}_{n-1})$ has the model
  $\beta \cup \nu_0 \cup \ldots \nu_{n-1}$, where $\beta(q) =
  \beta_{n-1}(q_{n-1})$, for all $q \in Q$. By the definition of
  $\pi(u)$: \[\theta_n \equiv \bigwedge_{q_{n-1} \in
    \fv{\Bool}{\theta_{n-1}} \cap Q_{n-1}} \left(q_{n-1} \rightarrow
  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]\right)\]

@@

  (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
    interpolant for $\pi(u)$, the following entailments hold: \begin{compactitem}
    \item $\theta_0 \models I_1[Q_0/Q]$, 
      %
    \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
      %
    \item $I_{n+1}[Q_{n+1}/Q] \wedge \theta_{n+1} \models \bot$. 
    \end{compactitem}
    We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$
    by induction on $n \geq 0$. This is sufficient to conclude because
    $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
    a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
    \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we
    have $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv
    \theta_0[Q/Q_0] \models I_1$. For the induction step $n>0$, we
    compute:
    \[\begin{array}{lcl}
    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] & \equiv & \text{ def. of $\AbsPost{\A}$} \\
    \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] & \models \\
    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n & \models & \text{ ($\Asterisk$)} \\
    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n & \models & \text{ ind. hyp.} \\ 
    I_n[Q_n/Q] & \hfill\text{\qed}
    \end{array}\]
    We prove the final point ($\Asterisk$). If
    $\beta_n\cup\nu_{n-1}\cup\nu_n \models
    \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q]$,
    for some valuations $\beta_{n-1} : Q_{n-1} \rightarrow \booli$ and
    $\nu_i : \vec{x}_i \rightarrow \Data^\I$, for $i=n-1,n$, then we
    build a valuation $\beta_{n-1}$ such that $\beta_{n-1}\cup\beta_n
    \cup \nu_{n-1}\cup\nu_n \models \AbsPost{\A}(\iota,a_1\ldots
    a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$, as follows. For each
    occurrence of a subformula $\Delta(q,a_n)$ in
    $\AbsPost{\A}(\iota,a_1\ldots a_{n-1})$ we set
    $\beta_{n-1}(q_{n-1}) = \true$ if $\beta_n\cup\nu_{n-1}\cup\nu_n
    \models \Delta(q,a_n)[Q_n/Q]$ and $\beta_{n-1}(q_{n-1}) = \false$,
    otherwise. Since there are no negated occurrences of such
    subformulae, the definition of $\beta_{n-1}$ is consistent, and
    the check $\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
    \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$
    is straightforward. \qed}

@@

\begin{algorithm}[t!]
{\scriptsize\begin{algorithmic}[0]
\State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
over the alphabet $\Sigma$ of input events

\State {\bf output}: $\true$ if $L(\A)=\emptyset$ and a data word $w
\in L(\A)$ otherwise
\end{algorithmic}}

{\scriptsize\begin{algorithmic}[1]
  \State $\preds \leftarrow \set{\bot}$, $\worklist \leftarrow
  \tuple{\rootNode}$, $\Art = \tuple{N,E,\Lambda,\lhd}$, where $N \leftarrow
  \emptyset$, $E \leftarrow \emptyset$, $\Lambda \leftarrow
  \set{(\rootNode,\iota)}$ and $\lhd \leftarrow \emptyset$
